Explaining the reportedly over-massive black holes in early-type galaxies with intermediate-scale discs

### 2015-11-04 papers/ellicular/sabotage/ellicular.tex
wed

\citep{liller1966,michard1984,djorgovski1985,jedrzejewski1987,BenderMoellenhoff1987,
carter1987,capaccioli1987,capaccioli1988,stromstrom1978,bender1988,bender1990,

@@

Not surprisingly, it is well known that 

@@

(a concept pioneered by \citealt{liller1960})\footnote{\cite{liller1960} 
used ellipticity profiles to help identify elliptical and spiral galaxy type.} 

@@

and she classified them as ``ES'' galaxies\footnote{\cite{strom1977} and \cite{stromstrom1978} 
classified the intermediate ``ES'' galaxy type as ``E/S0'', see also \cite{thompson1976}.}.  

@@

and referred to them as ``disk-ellipticals'' (or ``disky ellipticals'', \citealt{simienmichard1990}) 
thereby more closely associating them with elliptical galaxies than lenticular galaxies.   

@@

The same is true when searching for pointy isophotes that are shaped by the combination of the spheroid and a near edge-on disc 
(e.g.~\citealt{carter1978,carter1987,jedrzejewski1987,BenderMoellenhoff1987}). 
Today, most early-type galaxies are classified as ``fast rotators\footnote{The term 
``fast rotator'' is of course not new for describing certain early-type galaxies (e.g.~\citealt{bertinstiavelli1984,lakedressler1986}).}'' 

@@

In general, stellar discs are intrinsically flat and close to circular \citep{Andersen2001,AndersenBershady2002}; 

@@

Indeed, \citeauthor{graham2001} (\citeyear{graham2001}, his Figure 7) warned about such erroneous fits to late-type galaxies. 
However, this has led to the rejection of many early-type galaxy decompositions 

@@

Finally, in Section \ref{sec:disc} we briefly discuss our results in terms of galaxy evolution 
and we present the location of the ES galaxies relative to other disc galaxies 
in a common classification scheme of Hubble type versus spheroid-to-disc ratio. }

@@

\section{Intermediate-scale disc galaxies}
\label{sec:gal}
Three examples of galaxies with intermediate-scale discs are Mrk 1216, NGC 1332, and NGC 3115. 
In the following Sections, we present a photometric analysis of these three galaxies, 
and we compare our results with the kinematical analysis available from the literature for Mrk 1216 and NGC 3115. 
For the galaxies NGC 1332 and NGC 3115, we used $3.6~\rm \mu m$ images obtained with the InfraRed Array Camera (IRAC) 
onboard the \emph{Spitzer Space Telescope}. 
For the galaxy Mrk 1216, we used an archived Hubble Space Telescope (\emph{HST}) image  
taken with the Wide Field Camera 3 (WFC3) and the near-infrared \emph{F160W} filter ($H$-band). 
Our galaxy decomposition technique is extensively described in Savorgnan \& Graham (2015).
Briefly, the galaxy images were background-subtracted, and masks for contaminating sources were created. 
The one-dimensional Point Spread Function (PSF) was characterized using a Gaussian profile for the \emph{HST} observation 
and a \cite{moffat1969} profile for the \emph{Spitzer} observations.
We performed an isophotal analysis of the galaxies using the IRAF\footnote{IRAF 
is the Image Reduction and Analysis Facility, distributed by the National Optical Astronomy Observatory, 
which is operated by the Association of Universities for Research in Astronomy (AURA) 
under cooperative agreement with the National Science Foundation.} task {\tt ellipse}\footnote{Our analysis 
was performed before {\tt isofit} \citep{ciambur2015} was conceived or available. 
After {\tt isofit} was recently developed and implemented in IRAF, 
we employed it to re-extract the surface brightness profiles of the galaxies NGC 1332 and NGC 3115. 
We then repeated the analysis and checked that this change does not significantly alter our results. 
In fact, although {\tt isofit} provides a more accurate description of the isophotes in the presence of an inclined disc, 
the discs of NGC 1332 and NGC 3115 are relatively faint compared to the spheroidal components, 
therefore the differences between the light profile obtained with {\tt ellipse} and that obtained with {\tt isofit} 
are small for these two galaxies. } 
\citep{taskellipse}. 
The galaxy isophotes were modelled with a series of concentric ellipses, 
allowing the ellipticity, the position angle and the amplitude of the fourth harmonic to vary with radius.  
The decomposition of the surface brightness profiles was performed with software written by G. Savorgnan 
and described in Savorgnan \& Graham (2015).
We modelled the light profiles with a combination of PSF-convolved analytic functions, 
using one function per galaxy component.

\subsection{NGC 3115}
The presence of a disc in the {\bf central fast rotator} NGC 3115 
{\bf (e.g.~\citealt{strom1977,nieto1988,scorzabender1995}) }
is obvious due to its edge-on orientation (Figure \ref{fig:n3115}). 
Less obvious is the radial extent of this disc, if one only relies on a visual inspection of the galaxy image. 
The ellipticity profile (Figure \ref{fig:n3115}) is consistent with the presence of an intermediate-scale disc. 
Moreover, the kinematics of NGC 3115 \citep{arnold2011n3115} also disprove the presence of a large-scale disc, 
because the galaxy is rapidly rotating only within two galaxy half-light radii ($\sim 2 \times 50~\rm arcsec$), 
and the rotation significantly drops at larger radii.  
The unsharp mask of NGC 3115 (Figure \ref{fig:n3115}) betrays the presence of a faint edge-on nuclear ring, 
which can also be spotted as a small peak in the ellipticity profile 
(at semi-major axis length $R_{\rm maj} \sim 15~\rm arcsec$). 
Such rings are common in early-type galaxies \citep{michardmarchal1993}.
The spheroidal component of NGC 3115 is well described with a \cite{sersic1963} profile.
The highly inclined intermediate-scale disc is better fit with an $n<1$ S\'ersic profile 
(the S\'ersic index $n$ regulates the curvature of the S\'ersic profile) 
rather than with an exponential function, 
as explained by \cite{pastrav2013a}. 
The nuclear ring is modelled with a Gaussian function. \\
In comparison, \cite{lasker2014data} fit NGC 3115 with a bulge + disc + envelope, 
and measured a bulge half-light radius of $3.9~\rm arcsec$ and a bulge-to-total ratio of $0.12$. 
We describe this galaxy using a spheroid + intermediate-scale disc + nuclear ring, 
and obtain a spheroid half-light radius of $43.6~\rm arcsec$ and a spheroid-to-total ratio of $0.85$. 
We have used both kinematical information and ellipticity profiles, 
together with the surface brightness profile, 
to obtain a physically consistent and meaningful model.

@@

Acknowledging the correct structure of galaxies with intermediate-scale discs is important to properly understand their origin. 
According to the current paradigm of cosmological structure evolution, 
the genesis of massive early-type galaxies is characterized by two distinct phases: ``in-situ'' and ``ex-situ''. 
The first phase takes place in a young Universe (within its first $4~\rm Gyr$), 
when cold gas inflows produced short and intense bursts of star formation 
that created compact and dense conglomerates of stars with high velocity dispersion (e.g.~\citealt{prieto2013}). 
These naked and compact conglomerates, named ``red nuggets'' \citep{damjanov2009}, 
have been observed at high-redshift with half-light sizes of 1--2 kpc \citep{daddi2005,trujillo2006,vandokkum2008}.
In the second phase (last $10~\rm Gyr$), discs and stellar envelopes 
were accreted around these primordial conglomerates and the external parts of today's galaxies assembled on scales of 2--20 kpc 
(e.g.~\citealt{driver2013}). \\
Today's Universe is populated by an abundance of compact, massive spheroids, 
with the same physical properties -- mass and compactness -- as the high-redshift red nuggets \citep{graham2013review,GDS2015}. 
Some of these local compact massive spheroids are encased within a large-scale disc, 
that is to say they are the bulges of some lenticular and spiral galaxies.  
Over the last $10~\rm Gyr$ their spheroids have evolved by growing a relatively flat disc (e.g.~\citealt{pichon2011,danovich2012,stewart2013})
-- rather than a three-dimensional envelope -- 
which has increased the galaxy size but preserved the bulge compactness. 
{\bf Of course some lenticular/ES galaxies may have been built from mergers (e.g.~\citealt{querejeta2015}, and references therein). }
The other compact massive spheroids of today's Universe belong to some galaxies with intermediate-scale discs. 
Indeed, Mrk 1216, NGC 1271, NGC 1277, NGC 1332, and NGC 3115 are all local compact intermediate-scale disc galaxies 
with purely old ($>10~\rm Gyr$) stellar populations. 
These galaxies have undergone the lowest degree of disc growth. \\
In addition to the observational clues as to the actual physical components in galaxies with intermediate-scale discs, 
one can reason on other grounds as to why these compact galaxies are not comprised of an inner bulge 
plus large-scale disc plus outer envelope. 
If they were such three-component systems, then one would have two possibilities. 
The first possibility is that these galaxies were already fully assembled $10~\rm Gyr$ ago; 
this would explain their old stellar populations, 
but it would also imply that their discs and envelopes had already formed during the first $4~\rm Gyr$ of the Universe, 
in disagreement with the current cosmological picture. 
The second possibility is that only their inner bulges (with sizes of 0.1--0.2 kpc, 
according to past decompositions) originated in the first $4~\rm Gyr$ 
and they subsequently accreted a substantial disc and envelope. 
If this was correct, then we would observe high-redshift, star-like, naked bulges with stellar masses 
within a factor of a few times the currently observed red nuggets but sizes which are $10$ times smaller. 
However, a dramatically different expectation is reached 
if one considers these galaxies today as spheroid-dominated systems with an intermediate-scale disc; 
in this case, both the galaxy size and the spheroid size are compact (1--2 kpc). 
This implies that, among the local descendants of the high-redshift red nuggets, 
the compact intermediate-scale disc galaxies have undergone the lowest degree of disc growth. 
That is, the bulk of a compact intermediate-scale disc galaxy quickly assembled ``in-situ'' in a very young Universe 
and experienced very little evolution over the last $10~\rm Gyr$.

@@

\subsection{ES galaxies in the Hubble grid}
{\bf The existence of intermediate-scale discs has led to the suggestion of a likely continuum of disc sizes in early-type galaxies 
\citep{simiendevaucouleurs1986,capaccioli1988,bender1989,simienmichard1990,scorza1993,scorzavandenBosch1998} 
as opposed to a dichotomy of say nuclear discs versus large-scale discs.  
The presence of these intermediate-scale discs blurs the distinction between elliptical (E) and lenticular (S0) galaxies, 
or slow-rotators (SR) versus fast-rotators (FR), 
see \citeauthor{cappellari2011kmdr} (\citeyear{cappellari2011kmdr}, their Figure 1 and 2).   
There is a need for more than two bins (i.e.~nuclear vs.~large-scale disc, E vs.~S0, SR vs.~FR), 
and a need for a continuum in the (Hubble-Jeans)-like classification 
schemes\footnote{As noted by \cite{vandenbergh1997} and \cite{sandage2005}, 
it was Sir James \cite{jeans1928} who introduced the (tuning fork)-shaped diagram that encapsulated \citeauthor{Hubble1926}'s (\citeyear{Hubble1926}) 
elliptical-spiral sequence -- itself motivated by the earlier work of \cite{Jeans1919}.}.     \\
It has been argued that the classification scheme for ``elliptical'' galaxies should not be 
their apparent axis ratio as seen on the plane of the sky (originally E0--E7, see \citealt{hubble1936,sandage1961}, 
then E0--E4, see \citealt{liller1966} and also \citealt{gorbachev1970}), 
because this depends on the viewing angle rather than being intrinsic to the galaxy.  
Obviously when dealing with intermediate-scale discs, 
the boxy/discy shape of the isophotes is also a function of both radius and disc inclination, 
making single isophotal shape parameters problematic and particularly inappropriate for the ES class.  
Section 4 of \cite{kormendybender1996} points out additional complications in regard to this.  
Instead, early-type galaxies would be better quantified by their spheroid-to-total flux ratio, 
with a continuum from pure elliptical galaxies to disc-dominated lenticular galaxies (e.g.~\citealt{capaccioli1988}, and references therein).   
This spheroid-to-total flux (or better mass) ratio is widely recognised as an important quantity, 
and its broad range observed in the lenticular galaxies led \cite{cappellari2011kmdr} to present the ``Hubble comb''.   
Building on \citeauthor{vandenbergh1976}'s (\citeyear{vandenbergh1976}) diagram in which both late-type galaxies (i.e.~spiral galaxies) 
and early-type galaxies form separate prongs of an expanded Hubble-Jeans tuning fork, 
with an additional prong for disc galaxies hosting anaemic spiral patterns, 
\citeauthor{cappellari2011kmdr} (\citeyear{cappellari2011kmdr}, see also \citealt{kormendybender2012}) 
presented a scheme in which the morphological type (a, b, c, etc.) 
changed if the spheroid-to-total ratio changed.  
However the primary criteria used to assign the spiral galaxy type over the last half century \citep{sandage1961} 
has been the nature of the spiral arms -- the extent to which they are unwound, pitch angle, 
and the degree of resolution -- rather than the spheroid-to-total ratio.   
Indeed, in the Hubble Atlas of Galaxies, \cite{sandage1961} notes that Sa type galaxies exist with both small and large bulges 
(see also \citealt{VorontsovvelYaminovArkhipova1962} and \citealt{Vorontsovvelyaminovnoskova1973}). \\
The grid seen in Figure \ref{fig:grid} and discussed in \cite{graham2014review} 
was used by \citeauthor{freeman1970} (\citeyear{freeman1970}, his Figure 9), 
\cite{boroson1981}, \cite{kent1985}, \cite{kodaira1986}, \cite{simiendevaucouleurs1986} and others, 
and allows for the varying spheroid-to-total ratio among each galaxy type.   
Rather than using the extended classification scheme for early-type galaxies (E -- E$^+$ -- L$^-$ -- L$^o$ -- L$^+$ -- S0, 
corresponding to the numerical T-type ranging from -5 to 0, e.g.~\citealt{devaucouleurs1964,heidmann1972}), 
we have simply grouped them all together.  
While, collectively, spiral galaxies typically display a range of spheroid-to-total flux 
ratios from $0$ to $\approx 3/4$ (e.g.~\citealt{grahamworley2008}, and references therein), 
lenticular galaxies commonly display a range from $\approx 3/4$ to $\approx 0.1$ \citep{laurikainen2010}, and the ES class have larger ratios.  
The purpose here is not to provide a complete classification scheme, 
but to simply show where the early-type galaxies with intermediate-scale discs (ES) reside relative to the other 
Hubble galaxy types in this familiar diagram which has tended not to include the elliptical galaxies to date. }

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Hubblegrid.pdf}
\caption{The Hubble grid showing the location of the ES galaxies, intermediate between
elliptical (E) and lenticular (S0) galaxies, relative to the spiral galaxies.  }
\label{fig:grid}
\end{center}
\end{figure}

>>>>

\citep{liller1966,stromstrom1978,michard1984,djorgovski1985,jedrzejewski1987,BenderMoellenhoff1987,
carter1987,capaccioli1987,capaccioli1988,bender1988,bender1990,

@@

It is well known that 

@@

@@

and she classified them as ``ES'' galaxies.  

@@

and referred to them as ``disk-ellipticals'' (or ``disky ellipticals'', \citealt{simienmichard1990}).   

@@

Today, most early-type galaxies are classified as ``fast rotators'' 

@@

In general, stellar discs are intrinsically flat and close to circular (e.g.~\citealt{Andersen2001,AndersenBershady2002}); 

@@

This has led to the rejection of many early-type galaxy decompositions 

@@

Finally, in Section \ref{sec:disc} we briefly discuss our results in terms of galaxy evolution. }

@@

\section{Intermediate-scale disc galaxies}
\label{sec:gal}
Three examples of galaxies with intermediate-scale discs are Mrk 1216, NGC 1332, and NGC 3115. 
In the following Section, we present a photometric analysis of these three galaxies, 
and we compare our results with the kinematical analysis available from the literature for Mrk 1216 and NGC 3115. 
For the galaxies NGC 1332 and NGC 3115, we used $3.6~\rm \mu m$ images obtained with the InfraRed Array Camera (IRAC) 
onboard the \emph{Spitzer Space Telescope}. 
For the galaxy Mrk 1216, we used an archived Hubble Space Telescope (\emph{HST}) image  
taken with the Wide Field Camera 3 (WFC3) and the near-infrared \emph{F160W} filter ($H$-band). 
Our galaxy decomposition technique is extensively described in Savorgnan \& Graham (2015).
Briefly, the galaxy images were background-subtracted, and masks for contaminating sources were created. 
The one-dimensional Point Spread Function (PSF) was characterized using a Gaussian profile for the \emph{HST} observation 
and a \cite{moffat1969} profile for the \emph{Spitzer} observations.
We performed an isophotal analysis of the galaxies using the IRAF\footnote{IRAF 
is the Image Reduction and Analysis Facility, distributed by the National Optical Astronomy Observatory, 
which is operated by the Association of Universities for Research in Astronomy (AURA) 
under cooperative agreement with the National Science Foundation.} task {\tt ellipse}\footnote{Our analysis 
was performed before {\tt isofit} \citep{ciambur2015} was conceived or available. 
After {\tt isofit} was recently developed and implemented in IRAF, 
we employed it to re-extract the surface brightness profiles of the galaxies NGC 1332 and NGC 3115. 
We then repeated the analysis and checked that this change does not significantly alter our results. 
In fact, although {\tt isofit} provides a more accurate description of the isophotes in the presence of an inclined disc, 
the discs of NGC 1332 and NGC 3115 are relatively faint compared to the spheroidal components, 
therefore the differences between the light profile obtained with {\tt ellipse} and that obtained with {\tt isofit} 
are small for these two galaxies. } 
\citep{taskellipse}. 
The galaxy isophotes were modelled with a series of concentric ellipses, 
allowing the ellipticity, the position angle and the amplitude of the fourth harmonic to vary with radius.  
The decomposition of the surface brightness profiles was performed with software written by G. Savorgnan 
and described in Savorgnan \& Graham (2015).
We modelled the light profiles with a combination of PSF-convolved analytic functions, 
using one function per galaxy component.

\subsection{NGC 3115}
The presence of a disc in the {\bf central fast rotator} NGC 3115 
{\bf (e.g.~\citealt{strom1977,nieto1988,scorzabender1995}) }
is obvious due to its edge-on orientation (Figure \ref{fig:n3115}). 
Less obvious is the radial extent of this disc if one only relies on a visual inspection of the galaxy image. 
The ellipticity profile (Figure \ref{fig:n3115}) is consistent with the presence of an intermediate-scale disc. 
Moreover, the kinematics of NGC 3115 \citep{arnold2011n3115} also disprove the presence of a large-scale disc, 
because the galaxy is rapidly rotating only within two galaxy half-light radii ($\sim 2 \times 50~\rm arcsec$), 
and the rotation significantly drops at larger radii.  
The unsharp mask of NGC 3115 (Figure \ref{fig:n3115}) betrays the presence of a faint edge-on nuclear ring, 
which can also be spotted as a small peak in the ellipticity profile 
(at semi-major axis length $R_{\rm maj} \sim 15~\rm arcsec$). 
Such rings are common in early-type galaxies (e.g.~\citealt{michardmarchal1993}).
The spheroidal component of NGC 3115 is well described with a \cite{sersic1963} profile.
The highly inclined intermediate-scale disc is better fit with an $n<1$ S\'ersic profile 
(the S\'ersic index $n$ regulates the curvature of the S\'ersic profile) 
rather than with an exponential function, 
as explained by \cite{pastrav2013a}. 
The nuclear ring is modelled with a Gaussian function. \\
In comparison, \cite{lasker2014data} fit NGC 3115 with a bulge + disc + envelope, 
and measured a bulge half-light radius of $3.9~\rm arcsec$ and a bulge-to-total ratio of $0.12$. 
We describe this galaxy using a spheroid + intermediate-scale disc + nuclear ring, 
and obtain a spheroid half-light radius of $43.6~\rm arcsec$ and a spheroid-to-total ratio of $0.85$. 
We have used both kinematical information and ellipticity profiles, 
together with the surface brightness profile, 
to obtain a physically consistent and meaningful model.

@@

Acknowledging the correct structure of galaxies with intermediate-scale discs is important to properly understand their origin. 
According to the current paradigm of cosmological structure evolution, 
the genesis of massive early-type galaxies is characterized by two distinct phases: ``in-situ'' and ``ex-situ''. 
The first phase takes place in a young Universe (within its first $4~\rm Gyr$), 
when cold gas inflows produced short and intense bursts of star formation 
that created compact and dense conglomerates of stars with high velocity dispersion (e.g.~\citealt{prieto2013}). 
These naked and compact conglomerates, named ``red nuggets'' \citep{damjanov2009}, 
have been observed at high-redshift with half-light sizes of 1--2 kpc \citep{daddi2005,trujillo2006,vandokkum2008}.
In the second phase (last $10~\rm Gyr$), discs and stellar envelopes 
were accreted around these primordial conglomerates and the external parts of today's galaxies assembled on scales of 2--20 kpc 
(e.g.~\citealt{driver2013}). \\
Today's Universe is populated by an abundance of compact, massive spheroids, 
with the same physical properties -- mass and compactness -- as the high-redshift red nuggets \citep{GDS2015}. 
Some of these local compact massive spheroids are encased within a large-scale disc, 
that is to say they are the bulges of some lenticular and spiral galaxies.  
Over the last $10~\rm Gyr$ their spheroids have evolved by growing a relatively flat disc (e.g.~\citealt{pichon2011,danovich2012,stewart2013})
-- rather than a three-dimensional envelope -- 
which has increased the galaxy size but preserved the bulge compactness. 
{\bf Of course some lenticular/ES galaxies may have been built from mergers (e.g.~\citealt{querejeta2015}, and references therein). }
The other compact massive spheroids of today's Universe belong to some galaxies with intermediate-scale discs. 
Indeed, Mrk 1216, NGC 1271, NGC 1277, NGC 1332, and NGC 3115 are all local compact intermediate-scale disc galaxies 
with purely old ($>10~\rm Gyr$) stellar populations. 
These galaxies have undergone the lowest degree of disc growth. \\
In addition to the observational clues as to the actual physical components in galaxies with intermediate-scale discs, 
one can reason on other grounds as to why these compact galaxies are not comprised of an inner bulge 
plus large-scale disc plus outer envelope. 
If they were such three-component systems, then one would have two possibilities. 
The first possibility is that these galaxies were already fully assembled $10~\rm Gyr$ ago; 
this would explain their old stellar populations, 
but it would also imply that their discs and envelopes had already formed during the first $4~\rm Gyr$ of the Universe, 
in disagreement with the current cosmological picture. 
The second possibility is that only their inner bulges (with sizes of 0.1--0.2 kpc, 
according to past decompositions) originated in the first $4~\rm Gyr$ 
and they subsequently accreted a substantial disc and envelope. 
If this was correct, then we would observe high-redshift, star-like, naked bulges with stellar masses 
within a factor of a few times the currently observed red nuggets but sizes which are $10$ times smaller. 
However, a dramatically different expectation is reached 
if one considers these galaxies today as spheroid-dominated systems with an intermediate-scale disc; 
in this case, both the galaxy size and the spheroid size are compact (1--2 kpc). 
This implies that, among the local descendants of the high-redshift red nuggets, 
the compact intermediate-scale disc galaxies have undergone the lowest degree of disc growth. 
That is, the bulk of a compact intermediate-scale disc galaxy quickly assembled ``in-situ'' in a very young Universe 
and experienced very little evolution over the last $10~\rm Gyr$.

@@
